### 2014-04-11 doc/ApTranscriptome_manuscript.Rmd
Version 0.2 of manuscript finished and sent to group

We apply this approach in a study of two closely related eastern North American ant species, *Aphaenogaster rudis* and *A. picea* [@umphrey1996], which have more southern and northern distributions, respectively. Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguezcabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006]. As ectothermic species, ants experience a wide range of temperature conditions. Throughout the year, ant colonies persist through freezings and hot summers, and at shorter time scales, individual workers can experience substantial thermal shock as they leave the buffered ant nest to forage for food. Thus, at the organismal level we expect that ants utilize multiple molecular pathways to tolerate thermal stress.

@@

Colonies of the *Aphaenogaster picea-rudis* complex were collected from Molly Bog, Vermont (University of Vermont Natural Areas; lat, lon) and Durham, North Carolina ( 36.037° N -78.874° W), designated as *ApVT* and *ApNC* respectively. The phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996]. Though preliminary work indicates that the northern part of the distribution (including *ApVT*) to be a distinct clade and possibly species (Bernice Bacon DeMarco, pers. communication), we will refer to these as distinct colonies rather than species. These colonies were maintained in the lab for 5 months prior to experimentation.

From each species, we haphazardly collected 12 ants at the same time on 12 days. Each day, the 12 ants were placed in glass tubes into a water bath (...) at one of 12 randomly-assigned temperatures, every 3.5°C between 0° and 38.5°C, for one hour. The minimum and maximum temperatures were selected based on preliminary work showing that these temperatures are close to the critical minimum and maximum temperatures for an *Aphaenogaster* from VT (A. Nguyen, unpublished data) . At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C. mRNA was extracted from three pooled ants using a DNeasy kit (Qiagen Inc; Valencia, CA). Ants were homogenized in Buffer ATL with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY) and the standard Qiagen protocol was followed from this point.

@@

For each species, the 12 samples were barcoded and sequenced in a single lane of 2 x 100bp paired-end reads on an Illumina HiSeq 1500 yielding 200 and 160 million reads for the ApVT and ArNC samples respectively. Reads were filtered to remove Illumina adapter sequences and low quality bases using the program Trimmomatic [@lohse2012].

We assembled the sequenced reads into the full set of mRNA transcripts (the `transcriptome`) for the combined data set from both colonies using the Trinity *de novo* transcriptome assembly program [@grabherr2011]. *De novo* transcriptome assembly is prone to falsely identifying alternative transcripts and identifying inaccurate transcripts that are chimeric (e.g. regions of two separate transcripts with enough overlap to assemble into a false, or chimeric, third transcript) [@yang2013] To account for this, we first ran program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length, and second ran the program `uclust` which cluster sequences completely contained within longer sequences at greater than 90% similarity (see Supplemental for exact parameters). We used liberal values (90% similarity) to merge transcripts to account for the fact that we were including sequencing data from two colonies of the same species, and wanted to avoid assembling orthologs individually. We performed subsequent analyses at the transcript level due to limitations of working with a system lacking a reference genome, but refer to these transcripts as genes for simplicity.

@@

We quantified expression of each gene using the program `Sailfish` [@patro2013]. `Sailfish` reports gene expression as transcripts per million (TPM) which is the preferred measure of gene expression as values are comparable among sequencing runs [@wagner2012]. In addition, `Sailfish` corrects for sequence composition bias and gene length. As preliminary examination of the data (see supplemental...) indicated that 7°C samples may have been mis-labeled, we omitted these data from the analysis. The expression values were highly correlated between colonies at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both colonies was justified (Supplemental).

@@

For each thermally-responsive gene, we predicted expression levels across the tested thermal range using the final linear model separately for each colony. We used the predicted gene expression values to partition the thermally-responsive genes into five expression categories for each colony; genes that had greatest expression at high (> 31°) temperatures (*High*), low (< 10°C) temperatures (*Low*), intermediate (10 - 30°C) temperatures (*Intermediate*), both high and low temperatures (*Bimodal*) or were not thermally-responsive in that colony (*NotResp*). For the *Bimodal* group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.

To describe the overall molecular toolkit for thermal tolerance in *Aphaenogaster*, we performed gene set enrichment analysis (GSEA) on the intersection of each expression category in the two colonies using the program `topGO` [@alexa2006; @alexa2010] with the `parentChild` algorithm [@grossmann2007] implemented in R [@r_core_team2013]. Briefly, this approach identifies GO terms that are overrepresented in the significant genes relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms.

### Colony-level comparisons

To gain greater insight on the genetic basis of thermal tolerance in *Aphaenogaster*, we performed comparative analyses between the two colonies focusing on the thermally-responsive genes that had a significant colony x temperature interaction. For each gene, we predicted expression levels across temperatures for each colony using the full linear model, and then grouped genes into the four responsive categories, *High*, *Low*, *Intermediate*, and *Bimodal*, as well as a fifth category *Not Expressed*. To focus on differences in gene expression between the two colonies, we performed GSEA for the set difference of each category.

With the same set of genes, we explored the extent to which differences in the thermal reactionome between the colonies were due to changes in the mean or shape of the reaction norm for each individual gene. Following the example of Murren et al [-@murren2014], we tested for overall differences in the (a) mean expression value across temperatures (*M*), (b) slope of expression (*S*), (c) curvature of expression (*C*) and (d) wiggle of expression (*W*), which consists of all higher-order differences in shape not capture by the first three measures (formulas given in Appendix). As genes differed by five orders of magnitude in expression levels, we calculated the mean-standardized difference between the colonies $\Delta M = (Mean_{ApVT} - Mean_{Ar}) / \bar M$, where $\bar M = (Mean_{ApVT} + Mean_{Ar}) / 2$ yielding $\Delta M$, $\Delta S$, $\Delta C$, and $\Delta W$ for the mean, slope, curvature and wiggle of each gene, respectively. Using the mean-standardized values, we performed one-sample two-sided *t*-tests to determine if the overall mean, slope or curvature were greater in *ApVT* than *ApNC*.

@@

For the subset of genes that responded to temperature in one colony (i.e. *High* expression in the *ApVT* colony)  but not the other, we hypothesized that geographically-divergent selection may have favored constitutive expression. To test this idea, we compared levels of constitutive expression between *A22* and *ApNC* near the optimum temperature of 19.5° C . Specifically, we predicted higher constitutive expression in *ApNC* for genes with *High* expression in *A22*, and conversely, higher constitutive expression  in *A22* for *Low* genes in *ApNC* .

The *Intermediate* expressed genes are core molecular processes that are expressed at non-stressful temperatures, and shut-off when the organism experiences thermal stress. We hypothesized that if the more southern *ApNC* colony was more thermally-tolerant than *ApVT* colony, genes with 'Intermediate' expression (10-30° C) would be expressed across a wider range of temperatures. To test this with our data, for each *Intermediate* expressed gene in each colony, we randomly sampled 1,000 temperature values weighted by the expression function (e.g. point of maximum expression had highest probability of being sampled). We then calculated the standard deviation of this random draw, and performed a *t*-test comparing the standard deviations of expression between colonies.

In contrast, *Bimodal* expressed genes are those that are activated in response to thermal stress. Following the reasoning for *Intermediate* genes, we hypothesized that a more thermally-sensitive colony would activate these thermally-responsive genes earlier than a more thermally-tolerant colony. We calculated the standard deviation of expression as for *Intermediate* genes. The interpretation is different, as low standard deviation is a gene with nearly flat expression, while the maximum standard deviation of expression is for a highly-convex expression pattern. Thus, finding one colony had overall greater standard deviation of expression for *Bimodal* genes would be indicative of greater thermal sensitivity.

**Data and analysis availability**

@@

```{r expression_type_table, echo=FALSE, results='asis'}
pandoc.table(Ap.type.table, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each colony and their overlap.")
library(xtable)
xtable(Ap.type.table)
```

@@

We identified 22,495 genes with overall model fits of P < 0.05, retaining 9,809 after correcting for FDR. Of these transcripts, 8,044 significantly change expression with temperature including 6,727 that have a colony x temperature interaction. The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not expressed differed significantly between the colonies ($\chi^2 = 170, P < 0.0001$; Table 1). For both colonies, the majority of responsive transcripts were in the *Low* category (Fig. 1). More transcripts had *Intermediate* expression in *ApNC* than in *ApVT*, which had nearly twice as many bimodally-expressed transcripts  (Fig. 1, Table 1).

![Probability density function of peak expression for all thermally-responsive in *ApNC* (blue) and *ApVT* (red) colonies](../results/PDF_expression_all.png)

@@

In addition, we examined the reaction norms for genes that we *a priori* expected to have significant thermal responses. In particular, we looked at the expression patterns of heat shock proteins including ...

### Colony-level comparisons

We quantified differences in the reaction norms of thermally-responsive genes between colonies as changes to due shifts in mean ($\Delta M$), slope ($\Delta S$), curvature ($\Delta W$) and wiggle ($\Delta W$). Overall, we found that there was no significant difference in position of the mean (*t* = -0.95, *P* = 0.34), but changes in the shape due to slope (*t* = 24.67, *P* < 0.001), curvature (*t* = 15.71, *P* < 0.001) and wiggle (*t* = -12.45, *P* < 0.001) were all highly significant. On average, slope (95% CI 0.41 - 0.48) and curvature (95% CI 0.25 - 0.33) were greater in *ApVT* than *ApNC*, indicating a steeper response to temperature.

Of the total differences ($\Delta T$) in reaction norms between colonies for each gene, about 8% (95% CI 1 - 32%) was due to shifts in mean, 47% (95% CI 14 - 92%) due to shifts in slope, 39% (95% CI 0 - 67%) due to shifts in curvature and 0% (95% CI 0 - 87%) due to shifts in wiggle.

Plasticity is a well-characterized mechanism by which organism tolerate environmental change. Given our sampling of a warmer (*ApNC*) and colder (*ApVT*) climate species, we hypothesized that genes which are plastically-expressed under heat shock in the colder species would have greater constitutive expression in the warmer species, and vice versa. Consistent with one of our predictions, we found that constitutive expression of *High* genes from *ApVT* was greater in the more southern *ApNC* colony (Wilcoxon V = 911970, *P* < 0.001, 95% CI location shift -0.11 - 0.03). However, for genes with *Low* expression in *ApNC* we also found that constitutive expression was greater in *ApNC* than *ApVT* (Wilcoxon V = 2041149, *P* < 0.001, 95% CI location shift -0.23 - 0.15).

We tested for a transcriptome-wide pattern of thermal tolerance and sensitivity by examining the standard deviation of expression for the *Intermediate* expressed genes. Consistent with our expectations, the more southern *ApNC* colony had greater thermal breadth for *Intermediate* genes than the *ApVT* colony (*t*~df=1406~ = -9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64). In contrast, there was no difference in thermal sensitivity as measured by the standard deviation of expression for *Bimodal* expressed genes (*t*~df=1406~ = 0.31, *P* = 0.76).

We performed gene set enrichment analysis on the set difference of each expression category to reveal differences among the colonies response to thermal shock. For *High* genes, there were 4 enriched GO terms in *ApVT* and 12 in *ApNC*. For *Low* genes, there were 40 enriched GO terms in *ApVT* and 23 in *ApNC*. For *Intermediate* genes, there were 3 enriched GO terms in *ApVT* and 54 in *ApNC*. For *Bimodal* genes, there were 12 enriched GO terms in *ApVT* and 10 in *ApNC*.

@@

Thermal tolerance is a key trait for determining the survival of individual organisms, the persistence of populations and the geographic limits of species. While considerable research has focused on a few key genes (e.g. heat shock proteins) involved in thermal stress response [...], little is known about the functional diversity or transcriptome-wide patterns of thermal responsiveness. In this study, we characterized the thermal reactionome of sister ant species, *Aphaenogaster picea* and *Aphaenogaster rudis*. We found that nearly 10% of the genes in this species show significant thermal responsiveness. Moreover, we found ... differences in patterns of expression among these two species.

Two functional groups of *a priori* interest showed patterns consistent with our expectations. Responsive genes annotated to "heat shock proteins" were up-regulated at high temperatures (>31 C) in the northern *A. picea* colony, but not up-regulated in the putatively more heat tolerant *A. rudis* colony (Fig. 3a). Conversely, genes involved in the biological process of dormancy were up-regulated at cold temperatures (< 10 C) in the southern *A. rudis* colony, but not *A. picea* (Fig. 3b). As these genes represent only a handful of the responsive genes, the expression patterns of all responsive genes can be searched and viewed interactively at [https://johnsg.shinyapps.io/ApRxN-shinyapp/].

![Figure 3. goes here]

Responsiveness ... a temperate ant species are thermally responsive. The majority of these genes (about 90%) are up-regulated at high or low temperatures. Discuss the 26 "stress response" genes

Previous studies have using *Drosophila* [@sorensen2005] and a temperate flesh fly [@teets2012-1] have implicated GO terms .... as responding to heat stress. In this study, we found .... of these GO terms to be enriched at high temperatures for *Aphaenogaster*. However, when we looked at the exact identify of the genes, we only found X overlapping genes.

@@

A key outcome of this study was developing the first genomic tools for the study of adaptive differentiation in *Aphaenogaster*. Our comparative results are limited to two populations, and an important future step will be expanding this work to examine differences in gene expression among a larger number of populations. Moreover, the potential for species to adapt to climate change will depend in part on their genetic variation for thermal tolerance. The quantitative genetic studies required to estimate this adaptive potential have been out of reach of ant biologists due to the long lifespan of colonies, and inability of ants to mate in the lab. However, future work could leverage genomic information to infer the relatedness among colonies in the absence of a known pedigree, enabling estimation of genetic variation within populations and their adaptive potential.

>>>>

We apply this approach in a study of two closely related eastern North American ant species, *Aphaenogaster rudis* and *A. picea* [@umphrey1996], which have more southern and northern distributions, respectively. Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguezcabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006]. As ectothermic species, ants experience a wide range of temperature conditions. Throughout the year, ant species persist through freezings and hot summers, and at shorter time scales, individual workers can experience substantial thermal shock as they leave the buffered ant nest to forage for food. Thus, at the organismal level we expect that ants utilize multiple molecular pathways to tolerate thermal stress.

@@

*Aphaenogaster picea* was collected from Molly Bog, Vermont (University of Vermont Natural Areas; lat, lon) and *A. rudis* was collected from Durham, North Carolina ( 36.037° N -78.874° W), designated as *A. picea* and *A. rudis* henceforth. Though the phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], preliminary work supports that these species are separate species (Bernice Bacon DeMarco, pers. communication). species of these species were maintained in the lab for 5 months prior to experimentation.

From a single species of each species, we haphazardly collected 12 ants at the same time on 12 days. Each day, the 12 ants were placed in glass tubes into a water bath (...) at one of 12 randomly-assigned temperatures, every 3.5°C between 0° and 38.5°C, for one hour. The minimum and maximum temperatures were selected based on preliminary work showing that these temperatures are close to the critical minimum and maximum temperatures for an *Aphaenogaster* [@warren2013. At the end of the hour, the ants were flash frozen in liquid nitrogen and stored at -80°C. mRNA was extracted from three pooled ants using a DNeasy kit (Qiagen Inc; Valencia, CA). Ants were homogenized in Buffer ATL with zirconium silicate beads in a Bullet Blender (Next Advance; Averill Park, NY) and the standard Qiagen protocol was followed from this point.

@@

For each species, the 12 samples were barcoded and sequenced in a single lane of 2 x 100bp paired-end reads on an Illumina HiSeq 1500 yielding 200 and 160 million reads for the A. picea and A. rudis samples respectively. Reads were filtered to remove Illumina adapter sequences and low quality bases using the program Trimmomatic [@lohse2012].

We assembled the sequenced reads into the full set of mRNA transcripts (the `transcriptome`) for the combined data set from both species using the Trinity *de novo* transcriptome assembly program [@grabherr2011]. *De novo* transcriptome assembly is prone to falsely identifying alternative transcripts and identifying inaccurate transcripts that are chimeric (e.g. regions of two separate transcripts with enough overlap to assemble into a false, or chimeric, third transcript) [@yang2013] To account for this, we first ran program `CAP3` [@huang1999] to cluster sequences with greater than 90% similarity and merge transcripts with overlaps longer than 100 bp and 98% similar in length, and second ran the program `uclust` which cluster sequences completely contained within longer sequences at greater than 90% similarity (see Supplemental for exact parameters). We used liberal values (90% similarity) to merge transcripts to account for the fact that we were including sequencing data from two species of the same species, and wanted to avoid assembling orthologs individually. We performed subsequent analyses at the transcript level due to limitations of working with a system lacking a reference genome, but refer to these transcripts as genes for simplicity.

@@

We quantified expression of each gene using the program `Sailfish` [@patro2013]. `Sailfish` reports gene expression as transcripts per million (TPM) which is the preferred measure of gene expression as values are comparable among sequencing runs [@wagner2012]. In addition, `Sailfish` corrects for sequence composition bias and gene length. As preliminary examination of the data (see supplemental...) indicated that 7°C samples may have been mis-labeled, we omitted these data from the analysis. The expression values were highly correlated between species at each temperature treatment (r^2^ > 0.98) indicating that assembling the transcriptome with data from both species was justified (Supplemental).

@@

For each thermally-responsive gene, we predicted expression levels across the tested thermal range using the final linear model separately for each species. We used the predicted gene expression values to partition the thermally-responsive genes into five expression categories for each species; genes that had greatest expression at high (> 31°) temperatures (*High*), low (< 10°C) temperatures (*Low*), intermediate (10 - 30°C) temperatures (*Intermediate*), both high and low temperatures (*Bimodal*) or were not thermally-responsive in that species (*NotResp*). For the *Bimodal* group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C.

To describe the overall molecular toolkit for thermal tolerance in *Aphaenogaster*, we performed gene set enrichment analysis (GSEA) on the intersection of each expression category in the two species using the program `topGO` [@alexa2006; @alexa2010] with the `parentChild` algorithm [@grossmann2007] implemented in R [@r_core_team2013]. Briefly, this approach identifies GO terms that are overrepresented in the significant genes relative to all GO terms in the transcriptome, accounting for dependencies among the GO terms. To reduce redundancy in the enriched GO terms, we used the web program [ReviGO](http://revigo.irb.hr/).

### Species-level comparisons

To gain greater insight on the genetic basis of thermal tolerance in *Aphaenogaster*, we performed comparative analyses between the two species focusing on the thermally-responsive genes that had a significant species x temperature interaction. For each gene, we predicted expression levels across temperatures for each species using the full linear model, and then grouped genes into the four responsive categories, *High*, *Low*, *Intermediate*, and *Bimodal*, as well as a fifth category *Not Expressed*. To focus on differences in gene expression between the two species, we performed GSEA for the set difference of each category.

With the same set of genes, we explored the extent to which differences in the thermal reactionome between the species were due to changes in the mean or shape of the reaction norm for each individual gene. Following the example of Murren et al [-@murren2014], we tested for overall differences in the (a) mean expression value across temperatures (*M*), (b) slope of expression (*S*), (c) curvature of expression (*C*) and (d) wiggle of expression (*W*), which consists of all higher-order differences in shape not capture by the first three measures (formulas given in Appendix). As genes differed by five orders of magnitude in expression levels, we calculated the mean-standardized difference between the species $\Delta M = (Mean_{A. picea} - Mean_{Ar}) / \bar M$, where $\bar M = (Mean_{A. picea} + Mean_{Ar}) / 2$ yielding $\Delta M$, $\Delta S$, $\Delta C$, and $\Delta W$ for the mean, slope, curvature and wiggle of each gene, respectively. Using the mean-standardized values, we performed one-sample two-sided *t*-tests to determine if the overall mean, slope or curvature were greater in *A. picea* than *A. rudis*.

@@

To test if genes that were responsive to cold temperatures in the southern *A. rudis* species or hot temperatures in the northern *A. picea* species had higher constitutive expression in the contrasting species, we performed non-parametric two-sided Wilcoxon tests comparing expression values near the optimum of 19.5° C.

As critical maximum temperatures are more constrained than critical minimum temperatures [@warren2013; @hoffman2013], we predicted that the more northern *A. picea* species would have a broader thermal range of stable gene expression than *A. rudis*. To test this with our data, for each *Intermediate* expressed gene in each species, we randomly sampled 1,000 temperature values weighted by the expression function (e.g. point of maximum expression had highest probability of being sampled). We then calculated the standard deviation of this random draw, and performed a *t*-test comparing the standard deviations of expression between species. Conversely, we predicted that the thermal sensitivity of *Bimodal* expressed genes would be greater in *A. rudis* than *A. picea*. We performed the same analysis as with the *Intermediate* transcripts, but with low standard deviation being indicative of a relatively un-sensitive gene (nearly flat), while the maximum standard deviation of expression is for a highly-convex expression pattern. Thus, finding one species had overall greater standard deviation of expression for *Bimodal* genes would be indicative of greater thermal sensitivity.

### Data availability

@@

@@

We identified 22,495 genes with overall model fits of P < 0.05, retaining 9,809 after correcting for FDR. Of these transcripts, 8,044 significantly change expression with temperature including 6,727 that have a species x temperature interaction. The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not expressed differed significantly between the species ($\chi^2 = 170, P < 0.0001$; Table 1). For both species, the majority of responsive transcripts were in the *Low* category (Fig. 1). More transcripts had *Intermediate* expression in *A. rudis* than in *A. picea*, which had nearly twice as many bimodally-expressed transcripts  (Fig. 1, Table 1).

```{r expression_type_table, echo=FALSE, results='asis'}
pandoc.table(Ap.type.table, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate, both high and low (bimodal) temperatures or are not thermally-responsivefor each species and their overlap.")
```

@@

![Probability density function of peak expression for all thermally-responsive in *A. rudis* (blue) and *A. picea* (red) species](../results/PDF_expression_all.png)

As these genes represent only a handful of the responsive genes, the expression patterns of all responsive genes can be searched and viewed interactively at [https://johnsg.shinyapps.io/ApRxN-shinyapp/].

In addition, we examined the reaction norms for genes in the GO category "response to stress" which we *a priori* expected to have significant thermal responses....

### Species-level comparisons

We quantified differences in the reaction norms of thermally-responsive genes between species as changes to due shifts in mean ($\Delta M$), slope ($\Delta S$), curvature ($\Delta C$) and wiggle ($\Delta W$). Overall, we found that there was no significant difference in position of the mean (*t* = -0.95, *P* = 0.34), but changes in the shape due to slope (*t* = 24.67, *P* < 0.001), curvature (*t* = 15.71, *P* < 0.001) and wiggle (*t* = -12.45, *P* < 0.001) were all highly significant. On average, slope (95% CI 0.41 - 0.48) and curvature (95% CI 0.25 - 0.33) were greater in *A. picea* than *A. rudis*, indicating a steeper response to temperature.

Of the total differences ($\Delta T$) in reaction norms between species for each gene, about 8% (95% CI 1 - 32%) was due to shifts in mean, 47% (95% CI 14 - 92%) due to shifts in slope, 39% (95% CI 0 - 67%) due to shifts in curvature and 0% (95% CI 0 - 87%) due to shifts in wiggle.

Plasticity is a well-characterized mechanism by which organism tolerate environmental change. Given our sampling of a warmer (*A. rudis*) and colder (*A. picea*) climate species, we hypothesized that genes which are plastically-expressed under heat shock in the colder species would have greater constitutive expression in the warmer species, and vice versa. Consistent with one of our predictions, we found that constitutive expression was greater in the more southern *A. rudis* species for *High* genes in *A. picea* (Wilcoxon V = `r w1$statistic`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r w1$conf.int[1:2]`). However, for genes with *Low* expression in *A. rudis* we also found that constitutive expression was greater in *A. rudis* than *A. picea* (Wilcoxon V = `r w2$statistic`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r w2$conf.int[1:2]`).

We tested for a transcriptome-wide pattern of thermal tolerance and sensitivity by examining the thermal range) for the *Intermediate* expressed genes. Contrary to our expectations based on critical maximum and minimal temperatures, *A. rudis* had a transriptome-wide pattern of broader thermal range (*t*~df=1406~ = 9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64). In contrast, there was no difference in thermal sensitivity as measured by the standard deviation of expression for *Bimodal* expressed genes (*t*~df=1406~ = 0.31, *P* = 0.76).

We performed gene set enrichment analysis on the set difference of each expression category to reveal differences among the species response to thermal shock. For *High* genes, there were 4 enriched GO terms in *A. picea* and 12 in *A. rudis*. For *Low* genes, there were 40 enriched GO terms in *A. picea* and 23 in *A. rudis*. For *Intermediate* genes, there were 3 enriched GO terms in *A. picea* and 54 in *A. rudis*. For *Bimodal* genes, there were 12 enriched GO terms in *A. picea* and 10 in *A. rudis*.

![Expression (scaled to allow comparisons among genes) for genes in GO category "response to stress"](../results/GO_stress_High.png)

@@

Thermal tolerance is a key trait for determining the survival of individual organisms, the persistence of populations and the geographic limits of species. While considerable research has focused on a few key genes (e.g. heat shock proteins) involved in thermal stress response [...], little is known about the functional diversity or transcriptome-wide patterns of thermal responsiveness. In this study, we characterized the thermal reactionome of sister ant species, *Aphaenogaster picea* and *Aphaenogaster rudis*. We found that nearly 10% of the genes in this species show significant thermal responsiveness.

### Molecular toolkit of thermal responsiveness

While the sheer number of thermally-responsive transcripts makes generalizations difficult, groupings such as Gene Ontology (GO) can be used to identify categories associated with thermal stress. For heat response, the semantic space of the enriched GO terms included *gene expression*, *regeneration* and *cellular component organization or biogenesis*. ...previous work... For cold response, there were 65 enriched GO terms that grouped by semantic space into *protein modification process*, *sensory perception of mechanical stimulus*, *response to mechanical stimulus*, *folic acid metabolism*, *membrane lipid metabolism* and *protein complex biogenesis*. ...previous work...

*Bimodal* genes are general stress response genes that are activated under both heat and cold stress. The semantic space of the enriched GO terms included *regulation of metabolism*, *biological regulation*, *macromolecule biosynthesis*, ...

*Intermediate* genes represent non-essential molecular functions that are shut-down as conditions become stressful. The semantic space for the enriched GO terms included *lipid metabolic process*, *glycoprotein biosynthetic process*, *membrane lipid metabolic process*, *cellular protein metabolic process*, *membrane lipid biosynthetic process*. ...indicates shutting down of non essential molecule biosynthesis...

**Sara - comparisons of our gene list to other studies**

@@

### Transcriptome-wide comparisions

Of the 8,492 genes that were thermally-responsive, 80% differed in their responsiveness by colony. The majority of this difference in the reactionomes was due to changes in the slope ($\Delta S$) and curvature ($\Delta C$) which together accounted for nearly 90% of the differences between reaction norms for all transcripts. In a meta-analysis of phenotypic reaction norms, Murren et al. [-@murren2014] similarly found that differences in the shape of the reaction norms accounted for the majority of the differences. Thus, our finding suggests that this phenotypic pattern may be explained by greater genetic lability in the shape than the mean of gene expression. The relative role of plasticity versus adaptation in response to rapid environmental change is widely debated [@vellend...?]. Our data suggest that plasticity through changes in the responsiveness of gene expression are likely to occur more rapidly than changes in mean values.

An important note is that we would have completely missed this pattern if we had performed a differential expression study using only a few temperatures. Only by measuring gene expression across a thermal gradient and fitting a function to each transcript were we able to characterize the shape of the reaction norm of expression... To our knowledge, this is the first time that a reaction norm approach has been applied to the study of global gene expression.

Given the large differences in number (Table 1) and responsiveness of genes between the two species, a striking pattern is that 3-4x more genes are upregulated at cold than high temperatures. Intriguingly, this
result mirrors the observation in *Aphaenogaster* [@warren2013] and other ectotherms [@hoffmann2013] that critical maximum temperatures are more constrained than critical minimum temperatures. A plausible explanation is that organisms have fewer molecular mechanisms to tolerate heat than cold stress. Future studies ... implications for climate change as the evolutionary potential of ecototherms to climate change may be limited as suggested by ... [kingsolver?]

### Conclusions

A key outcome of this study was developing the first genomic tools for the study of adaptive differentiation in *Aphaenogaster*. Our comparative results are limited to two populations, and an important future step will be expanding this work to examine differences in gene expression among a larger number of populations. Moreover, the potential for species to adapt to climate change will depend in part on their genetic variation for thermal tolerance. The quantitative genetic studies required to estimate this adaptive potential have been out of reach of ant biologists due to the long lifespan of species, and inability of ants to mate in the lab. However, future work could leverage genomic information to infer the relatedness among species in the absence of a known pedigree, enabling estimation of genetic variation within populations and their adaptive potential.
